Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-09-03 aaai-2018/khodadadi-mcdonald-aaai19.tex
update paper

%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (Algorithms for Estimating Trends in Global Temperature Volatility)
%/Author (Khodadadi and McDonald)
/Author (John Doe, Jane Doe)}
\setcounter{secnumdepth}{2}

@@

@@

More recently, nonparametric models for both spatial and temporal data
have focused on using $\ell_1$-regularization for trend
estimation. \citet{KimKoh2009} proposed $\ell_1$-trend filtering for
univariate time series, which forms the basis of our methods. These
methods have been generalized to higher order temporal smoothness
~\cite{Tibshirani2014}, graph dependencies~\cite{WangSharpnack2016},
and, most recently, small, time-varying
graphs~\cite{HallacPark2017}. Our methodology is similar in flavor
to~\cite{HallacPark2017}, though it uses a different likelihood
function to emphasize variance estimation rather than trends in mean
signal. Furthermore, the focus is in high-dimensional, regular data
rather than a small number of changing graphs.

@@

@@

To further examine the performance of the proposed model, we next
compare it to three alternatives: a model which does not consider the
spatial smoothness (equivalent to fitting the model in
\autoref{sec:l1tf_var} to each time-series separately), a model which only imposes spatial smoothness, and a GARCH(1,1) model. We
simulated 100 datasets using the method explained above with $\sigma_s
\sim \mathsf{uniform}(4,7)$. The right panel of
\autoref{fig:true_fitted_var} shows the boxplot of the MAE for these
models. \attn{What is this next sentence trying to say?}
Interestingly, the proposed model with optimal parameters 
outperforms GARCH(1,1) in estimating the true value of the variance.

@@

\autoref{fig:cities_ts} shows the time-series of the 
temperature of three cities: Bloomington (USA), San Diego (USA) and
Manaus (Brazil). The time-series of Bloomington and San Diego show
clear cyclic behavior, though the amplitude of the cycles
changes. The time-series of Manaus does not
show any regular cyclic behavior. For this reason, we
first apply trend filtering to remove seasonal terms and de-trend
every time series. For each time-series, we found the optimal value of
the penalty parameter using \textit{k-fold cross-validation} with
$k=5$. We used the R package \textbf{genlasso} to perform these
computations \cite{arnold_efficient_2016}.  
\begin{figure}[tb]
	\centering
	\includegraphics[width=.2\textheight]{Figures/cities_ts}
 	\caption{Time-series of the temperature (in Kelvin) of three cities.}
 	\label{fig:cities_ts}
\end{figure} 
The blue curve in the left panel of Figure
\autoref{fig:bloom_estimatedSD} shows the time-series of the
temperature of Bloomington after detrending using this method.

\begin{figure}[tb]
  \centering
  \includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}
  \caption{Left: The residuals of the time-series of Bloomington
    (averaged weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Right: the same as middle panel but here the long horizon
    penalty is imposed. See the text for more details.} 
  \label{fig:bloom_estimatedSD}
\end{figure}

The red curve in the left panel of Figure \autoref{fig:bloom_estimatedSD}
shows the estimated SD (which is $\exp(h_t/2)$) of the residuals of
the time-series of Bloomington obtained from out proposed model. For
ease of analysis, we compute the average
of the estimated SD for each year. Both are shown in the middle panel of
\autoref{fig:bloom_estimatedSD}. To smooth the annual trend, we add
a long horizon penalty to \eqref{eq:l1tf_var}.
The estimated, smoothed SDs are shown
the right panel of \autoref{fig:bloom_estimatedSD}. The annual average
of the estimated SDs shows a linear trend with a positive slope.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=.8\columnwidth]{Figures/ts_estimatedVar}
  \caption{Detrended data and the estimated SD for
    a small midwestern city (left) and San Diego (right).\attn{Fix these}} 
  \label{fig:avg_change_estimatedSD}
\end{figure}

@@

\begin{align}
\nabla f^*(-D^\top v) = -\nabla_u f^*(u) D^\top ,\quad u=-D^\top v, \quad (\nabla_u f^*(u))_j=\log\bigg(\frac{y_j^2}{1-u_j}\bigg) \\
\nabla^2 f^*(-D^\top v)=D\nabla_u^2 f^*(u)D^\top, \quad (\nabla_u^2 f^*(u))_j=\mathbf{diag}\bigg(\frac{1}{1-u_j}\bigg)
\end{align}

@@

Figure \autoref{fig:cities_ts} shows the time-series of the
temperature of three cities: Bloomington (USA), San Diego (USA) and
Manaus (Brazil). The time-series of Bloomington and San Diego show
clear cyclic behavior. However, while it seems (qualitatively) that
these cycles can be modeled by a sinusoidal function for Bloomington,
the same is not true for San Diego. Also, the amplitude of the cycles
changes from some years to others. The time-series of Manaus does not
show any regular cyclic behavior. This demonstrates the first
difficulty in analyzing the variance of this data: to analyze the
variance, we first need to remove the cyclic terms from all
time-series. However, there is a lot of variations in the cyclic
behavior of the time-series of different locations. In addition, some
of these cycles cannot be easily modeled by a parametric function
\footnote{One might try to model the cycles by the summation of
  sinusoidal terms with different frequencies. However, for some
  time-series this may need many terms to be included in the summation
  to achieve a reasonable level of accuracy. In addition, this model
  cannot capture the non-stationarity in the cycles.}. To overcome
these issues, we use a non-parametric approach to remove the cyclic
terms from the time-series and de-trend them. This approach, called
\textit{$\ell_1$-trend filtering} is explained in Section 2 of the
text. We detrended each time-series separately using this method. For
each time-series, we found the optimal value of the penalty parameter
using \textit{k-fold cross-validation} with $k=5$. We used the R
package \textbf{genlasso} to perform these computations
\cite{arnold_efficient_2016}.

@@

The blue curve in the left panel of Figure \autoref{fig:bloom_estimatedSD} shows the time-series of the temperature of Bloomington after detrending using this method. This figure, reveals another difficulty in estimating the trend of volatility in this data: the variance of this signal, shows cyclic behavior. Also, the cycles are not regular and their amplitude and frequency change. Even if one can describe the behavior of the variance of the time-series at all locations using a single parametric model (for example a variant of the GARCH models \citep{bollerslev_generalized_1986}), it is not clear how the trend in the variance should be investigated in this framework. These observations motivate the need to develop a non-parametric framework for the problem at hand.

\begin{figure}[tb]
  \centering
  \includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}
  \caption{Left: The residuals of the time-series of Bloomington
    (averaged weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Right: the same as middle panel but here the long horizon
    penalty is imposed. See the text for more details.} 
  \label{fig:bloom_estimatedSD}
\end{figure}

The red curve in the left panel of Figure \autoref{fig:bloom_estimatedSD}
shows the estimated SD (which is $\exp(h_t/2)$) of the residuals of
the time-series of Bloomington obtained from out proposed model. To reduce the number of time-steps we work on the weekly averaged of the data. The curve of the estimated SD captures the periodic variations in the
SD of the signal. Just by looking at this curve, it is hard to say if
the SD is decreasing or increasing. Therefore, we compute the average
of the estimated SD for each year. The estimated SD together with this
annual average is shown in the middle panel of
\autoref{fig:bloom_estimatedSD}. As it can be seen, the annual trend
is not smooth. This is because in the optimization problem
\eqref{eq:l1tf_var}, the smoothness of the annual trend is not
encouraged. To remedy this, we add the following long horizon penalty to \eqref{eq:l1tf_var}:

>>>>

%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:

%\pdfinfo{
%	/Title (Algorithms for Estimating Trends in Global Temperature Volatility)
	%/Author (Khodadadi and McDonald)
%	/Author (John Doe, Jane Doe)}

\setcounter{secnumdepth}{2}

@@

\textcolor{red}{How is this first part related to the rest of the paper?}

@@

More recently, nonparametric models for both spatial and temporal data
have focused on using $\ell_1$-regularization for trend
estimation. \citet{KimKoh2009} proposed $\ell_1$-trend filtering for
univariate time series, which forms the basis of our methods. These
methods have been generalized to higher order temporal smoothness
~\cite{Tibshirani2014}, graph dependencies~\cite{WangSharpnack2016},
and, most recently, small, time-varying
graphs~\cite{HallacPark2017}.

\textcolor{red}{Our methodology is similar in flavor to~\cite{HallacPark2017}, with several fundamental differences. The main goal of Hallac et al. is to discover the time-varying structure of a network. To achieve this goal, they introduced two penalty terms into the objective function. One term penalizes the sum of the absolute values of the off-diagonal elements of the inverse covariance matrix. This term enforces sparsity on the estimated inverse covariance matrix and so helps in discovering which nodes in the network are connected. The other penalty term, penalizes differences in the estimated inverse covariance matrix from one time-step to the next. Our goal in the present work is to detect the temporal trend in the variance of each point in the network. The network in our problem is the grid of points over the earth and so the network's structure remains the same through time. The variance of each point, however changes. To capture these properties in our data using the objective function used in Hallac et al. (see Equation 2 in their paper), we need to have different temporal penalty terms for the diagonal and off-diagonal elements in the inverse covariance matrix. This results in unnecessary complication in the optimization problem. To overcome these difficulties, we introduced the new objective function \ref{eq:l1tf_var_st_mat}. This leads to a new optimization problem which we consider its solution next.}

though it uses a different likelihood
function to emphasize variance estimation rather than trends in mean
signal. Furthermore, the focus is in high-dimensional, regular data
rather than a small number of changing graphs.

@@

\attn{Arash: Can you be more specific?}

@@

To further examine the performance of the proposed model, we next
compare it to three alternatives: a model which does not consider the
spatial smoothness (equivalent to fitting the model in
\autoref{sec:l1tf_var} to each time-series separately), a model which only imposes spatial smoothness, and a GARCH(1,1) model. We
simulated 100 datasets using the method explained above with $\sigma_s
\sim \mathsf{uniform}(4,7)$. The right panel of
\autoref{fig:true_fitted_var} shows the boxplot of the MAE for these
models. \attn{What is this next sentence trying to say?}

\attn{that our model is better than GARCH in estimating the variance.}

Interestingly, the proposed model with optimal parameters 
outperforms GARCH(1,1) in estimating the true value of the variance.

@@

\autoref{fig:cities_ts} shows the time-series of the 
temperature of three cities: Indianapolis (USA), San Diego (USA) and
Manaus (Brazil). The time-series of Indianapolis and San Diego show
clear cyclic behavior, though the amplitude of the cycles
changes. The time-series of Manaus does not
show any regular cyclic behavior. For this reason, we
first apply trend filtering to remove seasonal terms and de-trend
every time series. For each time-series, we found the optimal value of
the penalty parameter using \textit{k-fold cross-validation} with
$k=5$. We used the R package \textbf{genlasso} to perform these
computations \cite{arnold_efficient_2016}.  
\begin{figure}[tb]
	\centering
	\includegraphics[width=.2\textheight]{Figures/cities_ts}
 	\caption{Time-series of the temperature (in Kelvin) of three cities.}
 	\label{fig:cities_ts}
\end{figure} 
The blue curve in the left panel of Figure
\autoref{fig:bloom_estimatedSD} shows the time-series of the
temperature of Indianapolis after detrending using this method.

\begin{figure}[tb]
  \centering
  \includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}
  \caption{Left: The residuals of the time-series of Indianapolis
    (averaged weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Right: the same as middle panel but here the long horizon
    penalty is imposed. See the text for more details.} 
  \label{fig:bloom_estimatedSD}
\end{figure}

The red curve in the left panel of Figure \autoref{fig:bloom_estimatedSD}
shows the estimated SD (which is $\exp(h_t/2)$) of the residuals of
the time-series of Indianapolis obtained from out proposed model. For
ease of analysis, we compute the average
of the estimated SD for each year. Both are shown in the middle panel of
\autoref{fig:bloom_estimatedSD}. To smooth the annual trend, we add
a long horizon penalty to \eqref{eq:l1tf_var}.
The estimated, smoothed SDs are shown
the right panel of \autoref{fig:bloom_estimatedSD}. The annual average
of the estimated SDs shows a linear trend with a positive slope.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=1\columnwidth]{Figures/ts_estimatedVar}
  \caption{Detrended data and the estimated SD for
    a small midwestern city (left) and San Diego (right).\attn{Fix these}} 
  \label{fig:avg_change_estimatedSD}
\end{figure}

@@

\begin{align}
& \nabla f^*(-D^\top v) =  -\nabla_u f^*(u) D^\top ,\\
& \quad u=-D^\top v, \quad (\nabla_u f^*(u))_j=\log\bigg(\frac{y_j^2}{1-u_j}\bigg) \\
& \nabla^2 f^*(-D^\top v)=D\nabla_u^2 f^*(u)D^\top, \quad (\nabla_u^2 f^*(u))_j=\mathbf{diag}\bigg(\frac{1}{1-u_j}\bigg)
\end{align}

@@

Figure \autoref{fig:cities_ts} shows the time-series of the
temperature of three cities: Indianapolis (USA), San Diego (USA) and
Manaus (Brazil). The time-series of Indianapolis and San Diego show
clear cyclic behavior. However, while it seems (qualitatively) that
these cycles can be modeled by a sinusoidal function for Indianapolis,
the same is not true for San Diego. Also, the amplitude of the cycles
changes from some years to others. The time-series of Manaus does not
show any regular cyclic behavior. This demonstrates the first
difficulty in analyzing the variance of this data: to analyze the
variance, we first need to remove the cyclic terms from all
time-series. However, there is a lot of variations in the cyclic
behavior of the time-series of different locations. In addition, some
of these cycles cannot be easily modeled by a parametric function
\footnote{One might try to model the cycles by the summation of
  sinusoidal terms with different frequencies. However, for some
  time-series this may need many terms to be included in the summation
  to achieve a reasonable level of accuracy. In addition, this model
  cannot capture the non-stationarity in the cycles.}. To overcome
these issues, we use a non-parametric approach to remove the cyclic
terms from the time-series and de-trend them. This approach, called
\textit{$\ell_1$-trend filtering} is explained in Section 2 of the
text. We detrended each time-series separately using this method. For
each time-series, we found the optimal value of the penalty parameter
using \textit{k-fold cross-validation} with $k=5$. We used the R
package \textbf{genlasso} to perform these computations
\cite{arnold_efficient_2016}.

@@

The blue curve in the left panel of Figure \autoref{fig:bloom_estimatedSD} shows the time-series of the temperature of Indianapolis after detrending using this method. This figure, reveals another difficulty in estimating the trend of volatility in this data: the variance of this signal, shows cyclic behavior. Also, the cycles are not regular and their amplitude and frequency change. Even if one can describe the behavior of the variance of the time-series at all locations using a single parametric model (for example a variant of the GARCH models \citep{bollerslev_generalized_1986}), it is not clear how the trend in the variance should be investigated in this framework. These observations motivate the need to develop a non-parametric framework for the problem at hand.

\begin{figure}[tb]
  \centering
  \includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}
  \caption{Left: The residuals of the time-series of Indianapolis
    (averaged weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Right: the same as middle panel but here the long horizon
    penalty is imposed. See the text for more details.} 
  \label{fig:bloom_estimatedSD}
\end{figure}

The red curve in the left panel of Figure \autoref{fig:bloom_estimatedSD}
shows the estimated SD (which is $\exp(h_t/2)$) of the residuals of
the time-series of Indianapolis obtained from out proposed model. To reduce the number of time-steps we work on the weekly averaged of the data. The curve of the estimated SD captures the periodic variations in the
SD of the signal. Just by looking at this curve, it is hard to say if
the SD is decreasing or increasing. Therefore, we compute the average
of the estimated SD for each year. The estimated SD together with this
annual average is shown in the middle panel of
\autoref{fig:bloom_estimatedSD}. As it can be seen, the annual trend
is not smooth. This is because in the optimization problem
\eqref{eq:l1tf_var}, the smoothness of the annual trend is not
encouraged. To remedy this, we add the following long horizon penalty to \eqref{eq:l1tf_var}:
